Generating Correlated Data for Omics Simulation

### 2025-05-08 manuscript/paper.qmd
Add benchmark of runtime + memory usage

Our methods are based off the well-known Gaussian copula method, which uses a multivariate normal distribution to capture aspects of the gene-gene dependence, without necessarily assuming that the data is normally distributed.
This multivariate normal distribution is parameterized by a $p \times p$ covariance matrix, $\Sigma$.
In omics, choosing this $\Sigma$ presents a challenge since $n < p$.
The most obvious choice is to use the sample covariance matrix $\widehat \Sigma$ of the reference data set (possibly after some transformation of the data).
However, using $\widehat Sigma$ is a poor choice for $\Sigma$.
Indeed, we consider a choice of $\Sigma$ to be successful if data sets of $n$ samples simulated using $\Sigma$ have sample covaraince matrices that are similar to $\widehat \Sigma$.
Counter-intuitively, data simulated using $\Sigma := \widehat \Sigma$ have sample covariance matrices that systematically and substantially differ from $\widehat \Sigma$, see @fig-intro.

@@

This is due to the fact that $\widehat \Sigma$ is low rank, specifically at most rank $n - 1$, which is less than $p$ in typical omics studies.
This limitation means that any data generated from the multivariate normal distribution of that $\Sigma$ will lie on a $n-1$-dimensional hyperplane.
Therefore, newly simulated samples will actually be linear combination of the samples of the reference data set.
Intuitively, this means that the entire probability distribution of possible $p \times n$ dimensional omics data sets have been squished into a $n-1$ dimensional space, generating much higher gene-gene correlations than in the reference data set.
In contrast, we know that omics contains at least some level of random noise that is approximately independent from one gene to the next.
For example, RNA-seq involves a random sampling of fragments from the true sample of RNA molecules and other omics modes involve various background noise effects.
We therefore need to "inject" some level of additional randomness that is independent between genes on top of the correlations captured by $\widehat \Sigma$

We present three methods - called here PCA, spiked Wishart, and corpcor - for making this choice of $\Sigma$, all which take the form
$$ \Sigma = D^2 + PP^T $$
where $D$ is a diagonal matrix capturing the injected independence and $PP^T$ is a low-rank correlation matrix derived from $\widehat \Sigma$.
An additional advantage of this special form is that data may be generated very efficiently, see Methods, compared to arbitrary $\Sigma$ which typically require a Cholesky or eigenvalue decomposition.

@@

The R package [SPsimSeq](https://github.com/CenterForStatistics-UGent/SPsimSeq) [@Assefa2020] provides a dedicated RNA-seq and single-cell RNA-seq simulator using a Gaussian copula approach to simulate gene dependence.
In contrast to this package, it uses a two-step randomization.
In short, it first fits distributions to each genes and then bins those distributions into discrete buckets.
Then, new data is generated by drawing from a multivariate normal distribution whose covariance matrix equals the sample covariance matrix of the reference data set.
These values are used to choose which bucket each gene is drawn from.
Then, each gene's final value is drawn uniformly and independently from within its chosen bucket.
Therefore, SPsimSeq induces correlation in the first multivarainte normal draw, but then weakens it in the second independent uniform step.
This two-step process makes it challenging to compare on theoretical grounds to our proposed methods since it will dependend upon parameters such as the number of buckets used.
More fine-grained buckets will induce higher correlation, and in the limit of infinite buckets, it should approach our PCA method with $k$ equal to the number of samples in the reference data set.
`SPsimSeq` is more specialized and full-featured for RNA-seq simulation, providing, for example, native differential expression (DE) options,.
In comparison, our `dependentsimr` package requires manually setting marginal expression values to inject DE, but also supports other marginal distributions for situations outside of RNA-seq.

>>>>

Our methods are based off the well-known Gaussian copula method, which uses a multivariate normal distribution to capture aspects of the gene-gene dependence, without necessarily assuming that the data is normally distributed.
This multivariate normal distribution is parameterized by a $p \times p$ covariance matrix, $\Sigma$.
In omics, choosing this $\Sigma$ presents a challenge since $n < p$.
The most obvious choice is to use the sample covariance matrix $\widehat \Sigma$ of the reference data set (possibly after some transformation of the data).
However, using $\widehat \Sigma$ is a poor choice for $\Sigma$.
Indeed, we consider a choice of $\Sigma$ to be successful if data sets of $n$ samples simulated using $\Sigma$ have sample covaraince matrices that are similar to $\widehat \Sigma$.
Counter-intuitively, data simulated using $\Sigma := \widehat \Sigma$ have sample covariance matrices that systematically and substantially differ from $\widehat \Sigma$, see @fig-intro.

@@

One reason this occurs is that $\widehat \Sigma$ is low rank, specifically at most rank $n - 1$, which is less than $p$ in typical omics studies.
This limitation means that any data generated from the multivariate normal distribution of that $\Sigma$ will lie on a $n-1$-dimensional hyperplane.
Therefore, newly simulated samples will actually be linear combination of the samples of the reference data set.
Intuitively, this means that the entire probability distribution of possible $p$ dimensional data have been squished into a $n-1$ dimensional space, generating much higher gene-gene correlations than in the reference data set.
In contrast, we know that omics contains at least some level of random noise that is approximately independent from one gene to the next.
For example, RNA-seq involves a random sampling of fragments from the true sample of RNA molecules and other omics modes involve various background noise effects.
We therefore need to "inject" some level of additional randomness that is independent between genes on top of the correlations captured by $\widehat \Sigma$

We present three methods - called here PCA, spiked Wishart, and corpcor - for making this choice of $\Sigma$, all which take the form
$$ \Sigma = D^2 + PP^T $$
where $D$ is a diagonal matrix capturing the injected independence and $PP^T$ is a low-rank correlation matrix derived from $\widehat \Sigma$.
An additional advantage of this special form is that data may be generated very efficiently, see Methods, compared to arbitrary $\Sigma$ which typically require a Cholesky or eigenvalue decomposition, see Supplemental Figure S3.

@@

The R package [SPsimSeq](https://github.com/CenterForStatistics-UGent/SPsimSeq) [@Assefa2020] provides a dedicated RNA-seq and single-cell RNA-seq simulator using a Gaussian copula approach to simulate gene dependence.
In contrast to this package, it uses a two-step randomization.
In short, it first fits distributions to each genes and then bins those distributions into discrete buckets.
Then, new data is generated by drawing from a multivariate normal distribution whose covariance matrix equals the sample covariance matrix of the reference data set.
These values are used to choose which bucket each gene is drawn from.
Then, each gene's final value is drawn uniformly and independently from within its chosen bucket.
Therefore, SPsimSeq induces correlation in the first multivarainte normal draw, but then injects additional independence in the second uniform step.
This two-step process makes it challenging to compare on theoretical grounds to our proposed methods since it will dependend upon parameters such as the number of buckets used.
More fine-grained buckets will induce higher correlation, and in the limit of infinite buckets, it should approach our PCA method with $k$ equal to the number of samples in the reference data set.
`SPsimSeq` is more specialized and full-featured for RNA-seq simulation, providing, for example, native differential expression (DE) options,.
In comparison, our `dependentsimr` package requires manually setting marginal expression values to inject DE, but also supports other marginal distributions for situations outside of RNA-seq.
